---
title: "Class11"
author: "Yassaman Fakhari (A18541145)"
format: html
toc: true
---

## Background

In this lab, we analyze AlphaFold structure prediction results. We will read the predicted PDB models, compare them with RMSD, examine confidence scores with pLDDT, analyze model flexibility with RMSF, inspect PAE scores, and look at sequence conservation.

```{r}
library(bio3d)
library(pheatmap)
library(jsonlite)
```

## Read AlphaFold results

First, unzip the AlphaFold results file.

```{r}
unzip("test_94b5b.result.zip", exdir = "test_94b5b")
```

Make a vector of input PDB file names that we can read into R.

```{r}
results_dir <- "test_94b5b"

pdb_files <- list.files(
  path = results_dir,
  pattern = "\\.pdb$",
  full.names = TRUE
)

basename(pdb_files)
```

Read all PDB models and superpose them.

```{r}
pdbs <- pdbaln(pdb_files, fit = TRUE, exefile = "msa")

pdbs
```

## RMSD analysis

RMSD measures structural distance between coordinate sets. We can calculate RMSD between all pairs of AlphaFold models.

```{r}
rd <- rmsd(pdbs, fit = TRUE)

range(rd)
```

## Make RMSD heatmap

```{r}
colnames(rd) <- paste0("m", 1:ncol(rd))
rownames(rd) <- paste0("m", 1:nrow(rd))

pheatmap(rd)
```

## Plot pLDDT scores

The AlphaFold PDB files store pLDDT confidence scores in the B-factor column. Higher pLDDT means higher confidence.

```{r}
plotb3(
  pdbs$b[1,],
  typ = "l",
  lwd = 2,
  ylab = "pLDDT Score",
  xlab = "Residue Position"
)

points(pdbs$b[2,], typ = "l", col = "red")
points(pdbs$b[3,], typ = "l", col = "blue")
points(pdbs$b[4,], typ = "l", col = "darkgreen")
points(pdbs$b[5,], typ = "l", col = "orange")
```

## Core fitting

Next, find the most consistent rigid core of the models and use this core for fitting.

```{r}
core <- core.find(pdbs)

core.inds <- print(core, vol = 0.5)

xyz <- pdbfit(
  pdbs,
  core.inds,
  outpath = "corefit_structures"
)
```

# RMSF analysis

RMSF measures conformational variation across residue positions.

```{r}
rf <- rmsf(xyz)

plot(
  rf,
  type = "l",
  ylab = "RMSF",
  xlab = "Residue Position"
)
```

## PAE analysis

AlphaFold also produces Predicted Aligned Error, or PAE, values in JSON files.

```{r}
pae_files <- list.files(
  path = results_dir,
  pattern = ".*model.*\\.json",
  full.names = TRUE
)

pae_files
```

Read the first and last model PAE files.

```{r}
pae1 <- read_json(pae_files[1], simplifyVector = TRUE)
pae_last <- read_json(pae_files[length(pae_files)], simplifyVector = TRUE)

pae1$max_pae
pae_last$max_pae
```

## PAE plot for model 1

```{r}
plot.dmat(
  pae1$pae,
  xlab = "Residue Position (i)",
  ylab = "Residue Position (j)",
  grid.col = "black",
  zlim = c(0, 30)
)
```

# PAE plot for model 5

```{r}
plot.dmat(
  pae_last$pae,
  xlab = "Residue Position (i)",
  ylab = "Residue Position (j)",
  grid.col = "black",
  zlim = c(0, 30)
)
```

## Conservation analysis

```{r}
aln_file <- list.files(
  path = results_dir,
  pattern = "\\.a3m$",
  full.names = TRUE
)

aln_file
```

Read alignment

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)

dim(aln$ali)
```

Score residue conservation.

```{r}
sim <- conserv(aln)

m1.pdb <- read.pdb(pdb_files[1])

nres <- length(unique(m1.pdb$atom$resno))

plotb3(
  sim[1:nres],
  ylab = "Conservation Score",
  xlab = "Residue Position"
)
```

# Consensus sequence

```{r}
con <- consensus(aln, cutoff = 0.9)

con$seq
```

# Save conservation scores into occupancy column of model 1

```{r}
occ <- vec2resno(
  sim[1:nres],
  m1.pdb$atom$resno
)

write.pdb(
  m1.pdb,
  o = occ,
  file = "m1_conserv.pdb"
)
```

## Conclusion

The AlphaFold models were compared using RMSD, pLDDT, RMSF, PAE, and conservation analysis. The RMSD heatmap shows how similar or different the predicted models are from each other. The pLDDT plot shows confidence across residue positions. The RMSF plot highlights regions that vary more between models. The PAE plots show prediction error across residue pairs. The conservation analysis identifies more conserved positions in the sequence alignment.















