---
title: "Class14: RNASeq Mini Project"
author: "Yassaman Fakhari (PID A18541145)"
format: pdf
toc: true
---

## Background 

In today's mini project we will analyze data fromm GEO entry: GSE37704, which is associated with the following publication:

> Trapnell C, Hendrickson DG, Sauvageau M, Goff L et al. "Differential analysis of gene regulation at transcript resolution with RNA-seq". Nat Biotechnol 2013 Jan;31(1):46-53. PMID: 23222703

The authors report on differential analysis of lung fibroblasts in response to loss of the developmental transcription factor HOXA1.

## Data Import 

Read counts and metadata CSV files 

```{r}
library(DESeq2)
```

```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

colData = read.csv(metaFile, row.names=1)
head(colData)
```

```{r}
countData = read.csv(countFile, row.names=1)
head(countData)
```

Hmm... remember that we need the countData and colData files to match up so we will need to remove that odd first column in countData namely countData$length.

```{r}
countData <- as.matrix(countData[, -1])
head(countData)
```

```{r}
countData = countData[rowSums(countData) > 0, ]
head(countData)
```

### Sanity Check 

```{r}
metadata
```

## Setup DESeq Object 

```{r}
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ condition)

dds <- DESeq(dds)
dds
```

```{r}
res = results(dds)
summary(res)
```

## Run DESeq Analysis Pipeline 

## Extract the results 

Big table with log2fc and pvalues 

## Data Visualization 

volcano plot 

## Add Annotation Data 


add gene symbol and entrez ids

## Pathway Analysis

KEGG, GO and REACTOME

## Save Our Results 