---
title: "Class05: Data Viz with ggplot2"
author: "Yassaman Fakhari (PID A18541145)"
format: pdf
toc: true
---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *"base R" graphics* like the 'plot()' function and add on packages like **ggplot2**

Let's compare how we make a simple figure with these two systems:
We can use the in-built "cars" dataset:
```{r}
head(cars)
```

```{r}
plot(cars)
```


Before I can use ggplot2 I need to install it on my computer.
To do this we can use the function 'install.packages(ggplot2)'

Once installed we need to load up the package into our R brain:
```{r}
library(ggplot2)
```

The main function in the **ggplot** package is called ggplot()
```{r}
ggplot(cars)
```


every ggplot has at least 3 layers:
  - the **data** (a data.frame of the stuff we want to plot)
  - the **aes**thetics (how the data maps on the plot)
  - the **geom** layer (how you want the plot drawn, ex. points, lines, etc.)
```{r}
ggplot(cars, aes(x=speed, y=dist)) + 
  geom_point()
```

## Add Some Cutsom Features
Let's add a trend line that shows the relationsip between speed and distance.

```{r}
ggplot(cars, aes(x=speed, y=dist)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  theme_linedraw() +
  labs(title="Stop Dist of Old Cars",
       x="Speed (MPH)",
       y= "Distance (ft)")
```

> Q. Can you make the 'geom_smooth()' function produces a linear straight line fit to the data and turn off the 'grey' error region

```{r}
ggplot(cars, aes(x=speed, y=dist)) + 
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  theme_bw() +
  labs(title="Stop Dist of Old Cars",
       x="Speed (MPH)",
       y= "Distance (ft)")
```

## Gene Expression Figure

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
sum(genes$State == "up")
```

```{r}
table(genes$State)
```

My first plot attempt

```{r}
ggplot(genes) + 
  aes(x=Condition1, y=Condition2 ,col=State) +
  geom_point() +
  scale_color_manual(values=c("magenta", "gray", "turquoise")) +
  theme_bw() + 
  labs(x="No Drug", y="Drug", 
       title="Expression Changes Upon GLP-1 Inhibitor Treatment")
```

## Going Further 
Here we read the famous gapmider dataset:
```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many entries (ex. rows) are in this dataset?

```{r}
nrow(gapminder)
```

> Q. How many different "country" enteries are in this dataset?

```{r}
length(unique(gapminder$country))
```


Let's make our first plot of the entire dataset:
Plot of "gdpPercap" vs "lifeExp" colored by "continent"

```{r}
p <- ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha=0.3)
p
```

```{r}
p
p + 
  facet_wrap(~continent)
```


Make a plot for years 1977 and 2007 only (not all the years in the dataset).

> Q. First use the **dplyr** package and the 'filter()' function from that package to extract the rows from the year 2007

```{r, message=FALSE, warning=FALSE}
library(dplyr)
```

```{r}
g07 <- filter(gapminder, year == 2007)
g77 <- filter(gapminder, year == 1977)
g <- filter(gapminder, year == 2007 | year == 1997)
```

```{r}
ggplot(g) +
  aes(gdpPercap, lifeExp, col=continent, size=pop) +
  geom_point() +
  facet_wrap(~year)
```
> Q. Make a histogram of lifeExp colored by continent (try using 'fill=continent' or 'col=continent)

```{r}
ggplot(gapminder) +
  aes(lifeExp, fill=continent) +
  geom_histogram()
```

> Q. Make a histogram of lifeExp faceted by continent

```{r}
ggplot(gapminder) +
  aes(lifeExp) +
  geom_histogram() +
  facet_wrap(~continent)
```


