---
title: "Class07: Machine Learning 1"
author: "Yassaman Fakhari (PID:18541145)"
format: pdf
toc: true
---

##Background 

Today we will explore some core machine learning methods that are very popular in bioinformatics.These include **clustering** and **dimensionallity reduction**.

## K-Means Clustering 

The main function in "base" R for K-means clustering is called 'kmeans()'.

Before we go too deep let's make up some "simple" data that we can cluster and know if we are getting a good answer or not. To do this we can use 'rnorm()' function:

```{r}
hist( rnorm(10000, mean=3) )
```

```{r}
x <- c( rnorm(30, mean=-3), rnorm(30, +3) )
```

```{r}
z <- cbind(x=x, y=rev(x))
plot(z)
```

```{r}
p <- 1:5
rev(p)
```

Now we can run 'kmeans()' on this input 'z' and see what the results look like. 

```{r}
km <- kmeans(z, center=2)
km
```

```{r}
attributes(km)
```

> Q. How many pounts are in each cluster?

```{r}
km$size
```

> Q. What component of your result object details cluster assignment/membership?

```{r}
km$cluster
```


> Q. What component of your result object details cluster center?

```{r}
km$centers
```

> Q. Plot 'z' colored by the kmeans cluster assignment and add cluster centers as blue points.

```{r}
plot(z, col=c("magenta","turquoise"))
```

```{r}
plot(z, col=km$cluster)
points(km$centers, col="blue", pch=15)
```

> Q. Run a K-means clustering and plot the results asking for 4 clusters. (K=4)

```{r}
km2 <- kmeans(z, centers = 4)

plot(z, col = km2$cluster)
points(km2$centers, col = "purple", pch = 15)
```

> **N.B.** You need to tell L-means the number of clusters (i.e. set 'centers=2')!!

One approach is to try different values for 'centers' and then pick the best

```{r}
ans <- NULL
for(i in 1:10) {
  km <- kmeans(z, centers=i)
  ans <- c(ans, km$tot.withinss)
}

plot(ans,type="o", 
     xlab="Number of Clusters", ylab="Total Sum of Squars Distance")
```

## Hierarchical Clustering

The main function in "base" R for Hierarchical Clustering is called 'hclust()'.

This function does not take your "raw" data for clustering. You must first build a "distance matrix" from your data and pass this as a input to 'hclust()'.

```{r}
d <- dist(z)
hc <- hclust(d)
hc
```

There is a bespoke 'plot()' method for 'hclust()' result objects.

```{r}
plot(hc)
abline(h=8, col='red')
```

Once we have out 'hclust()' object (out"tree" of "cluster dendrogram") we can *"cut"* the tree to reveal the clustering pattern.

```{r}
cutree(hc, h=8)
```

```{r}
cutree(hc, k=4)
```

> Q. Make a plot of 'z' with your hclust results (i.e. colored by cluster membership)

```{r}
grps <- cutree(hc, k=2)
plot(z, col=grps)
```

## Principal Component Analysis (PCA)

PCA is a dimensionallity reduction method that is popular for revealing patterns in complex datasets. 

### Analysis of UK Food Data

Let's look at some data on the eating habits of folks from the UK to see if there are patterns and trends that have some regions being distinct form others. 

## Data Import

The data is made available in CSV format so we can use the 'read.csv()'.

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names = 1)
x
```

> Q.  How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

The number of rows and columns in the data frame can be determined using functions such as dim(), nrow(), and ncol().

```{r}
dim(x)
```

```{r}
head(x)
```


> Q. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer using read.csv(url, row.names = 1) because it is more concise and avoids the need for additional steps after importing the data. This approach is also less error-prone, since it directly assigns the first column as row names during the import process. In contrast, manually setting row names with row.names(x) <- x[,1] and then removing the column requires extra lines of code and increases the chance of mistakes (such as selecting the wrong column). However, the manual approach can be more flexible if the row names need to be chosen or modified after inspecting the dataset. Overall, the row.names = 1 method is more robust for standard cases, while the manual method is useful when more control is needed.

```{r}
head(x)
```

```{r}
barplot(as.matrix(x), beside = F, col = rainbow(nrow(x)))
```

## Tidy the Data

Fix anything that went wrong with the data import.

```{r}
library(tidyr)

# Convert data to long format for ggplot with the 'pivot_longer()'
x_long <- x |>
          tibble::rownames_to_column("Food") |>
          pivot_longer(cols = -Food,
                       names_to = "Country",
                       values_to = "Consumption")
dim(x_long)
```

```{r}
head(x_long)
```

# Create Grouped Bar Plot 

```{r}
library(ggplot2)
```

```{r}
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col() +
  theme_bw()
```

> Q5 We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagnoal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

```{r}
library(pheatmap)

pheatmap(as.matrix(x))
```

## Exploratory Analysis

Make some plots to help make sense of obvious trends...

> Q6 Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data set?

England, Wales, and Scotland cluster closely together, showing they have similar food consumption patterns. Northern Ireland is separated from the other three, indicating its diet is different. From the heatmap, Northern Ireland appears to have higher consumption of foods like fresh potatoes and soft drinks, which helps explain why it forms its own cluster. However, it is still not completely obvious which variables drive all the differences without further analysis.

> **Key-point**: Even relatively small datasets can prove challenging to interpret.

## PCA to the Rescue 

The main function in "base" R for PCA is called 'prcomp()'. This functions wants the 'observations' to be rows and the 'variables' to be columns.

So here we need to take the transpose of our 'x' input object.

```{r}
pca <- prcomp( t(x) )
summary(pca)
```

The returned 'pca' object has components that we can use to make our main result figures:

```{r}
attributes(pca)
```

The main result figure from this analysis is called a "PC score plot" or "ordenation plot", "PC plot" or "PC1 vs PC2 plot".

This plot shows how samples (in this case countries) relate to each other along our new PC axis.

This is our new "reduced-dimensional space". In this case 2 dimesions, PC1 and PC2, that capture most of the variance in the original 17 dimensional data-set.

```{r}
library(ggplot2)

ggplot(pca$x) +
  aes(PC1, PC2) +
  geom_point()
```

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names = 1)

dim(x)   # should be 17 4

pca <- prcomp(t(x))

df <- as.data.frame(pca$x)
df$Country <- rownames(df)

dim(df)  # should be 4 5

mycols <- c("orange","red","blue","darkgreen")

ggplot(df) +
  aes(PC1, PC2, label = Country) +
  geom_point(col = mycols, size = 3) +
  geom_text(col = mycols, size = 2, vjust = -0.5)
```

```{r}
ggplot(pca$x) + 
  aes(PC1, PC2, label=row.names(pca$x)) + 
  geom_point(col=mycols) + 
  geom_text(size=2, vjust=2, col=mycols)
```

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(row.names(pca$rotation), PC1)) +
  geom_col()
```

